---
title: "codeforproject"
author: "Katrine Munkholm"
date: "2026-06-03"
output: html_document
---

```{r}
pacman::p_load( "tidyverse",
  "lubridate",
  "httr2",
  "jsonlite",
  "rsample",
  "caret",
  "yardstick",
  "car",
  "ggtext",
  "scales")
```

#Read CSV

```{r}
Zdf <- read_csv2("data.csv")
```


#ENGINEERED VARIABLES
#Create column with journey ID for each unique journey, connecting individual records.
```{r}
Zdf <- Zdf %>%
  group_by(FORSTE_KOREDATO, TOGNUMMER, DELSTR_RETNING) %>%
  mutate(JOURNEY_ID = cur_group_id()) %>%
  ungroup()
```

#Create column for type of day. 1 = weekdays, 0 = weekends
```{r}
Zdf <- Zdf %>%
  mutate(
    FORSTE_KOREDATO = dmy(FORSTE_KOREDATO),
    WEEKDAY = if_else(
      wday(FORSTE_KOREDATO, week_start = 1) <= 5,
      1,
      0
    )
  )
```


#WEATHER SPECIFIC
#1 Create tribble for coordinates (For obtaining weather data)

```{r}
stations <- tribble(
  ~STAT_KODE, ~station_name, ~latitude,   ~longitude,
  "FA",       "Fredericia",  55.5682297,  9.7399931,
  "VJ",       "Vejle",       55.706389,   9.536944,
  "SD",       "Skanderborg", 56.043219,   9.924794,
  "AR",       "Aarhus",      56.150000,  10.204722
)
```

#Add coordinates to df by STAT_KODE
```{r}
Zdf <- Zdf %>%
  left_join(stations, by = "STAT_KODE")
```

#Create columns for correct date/time format for Open-Meteo to obtain weather data
```{r}
Zdf <- Zdf %>%
  mutate(
    PLANTID_datetime = lubridate::dmy_hm(
      PLANTID,
      tz = "Europe/Copenhagen"
    ),

    WEATHER_HOUR = format(
      lubridate::round_date(
        PLANTID_datetime,
        unit = "hour"
      ),
      "%Y-%m-%dT%H:%M"
    ),
    HOUR_OF_DAY = factor(
      format(PLANTID_datetime, "%H"),
      levels = sprintf("%02d", 0:23),
      ordered = TRUE
    )
  )
```

#Retrieve hourly weather information
```{r}
get_weather_openmeteo <- function(lat, lon, start_date, end_date, station_code) {
  
  hourly_vars <- c(
    "temperature_2m",
    "precipitation",
    "snowfall",
    "weather_code",
    "wind_speed_10m",
    "surface_pressure"
  )
  
  resp <- request("https://archive-api.open-meteo.com/v1/archive") %>%
    req_url_query(
      latitude = lat,
      longitude = lon,
      start_date = start_date,
      end_date = end_date,
      hourly = paste(hourly_vars, collapse = ","),
      timezone = "Europe/Copenhagen",
      models = "era5"
    ) %>%
    req_perform() %>%
    resp_body_string()
  
  parsed <- fromJSON(resp, simplifyDataFrame = TRUE)
  
  weather <- as_tibble(parsed$hourly)
  names(weather)[1] <- "WEATHER_HOUR"
  
  weather %>%
    mutate(
      WEATHER_HOUR = as.POSIXct(
  WEATHER_HOUR,
  format = "%Y-%m-%dT%H:%M",
  tz = "Europe/Copenhagen"
),
      STAT_KODE = station_code
    )
}

weather_df <- pmap_dfr(
  list(stations$latitude, stations$longitude, stations$STAT_KODE),
  ~ get_weather_openmeteo(..1, ..2, "2025-01-01", "2025-12-31", ..3)
)

weather_df <- weather_df %>%
  mutate(
    WEATHER_HOUR = format(WEATHER_HOUR, "%Y-%m-%dT%H:%M")
  )
```

#Mutate time to correct format
```{r}
weather_df <- weather_df %>%
  mutate(
    WEATHER_HOUR = as.POSIXct(
      WEATHER_HOUR,
      format = "%Y-%m-%dT%H:%M",
      tz = "Europe/Copenhagen"
    )
  )

Zdf <- Zdf %>%
  mutate(
    WEATHER_HOUR = as.POSIXct(
      WEATHER_HOUR,
      format = "%Y-%m-%dT%H:%M",
      tz = "Europe/Copenhagen"
    )
  )

#Check they are both "POSIXct" "POSIXt"
class(Zdf$WEATHER_HOUR)
class(weather_df$WEATHER_HOUR)
```

#Add the weather information to the main df
```{r}
Xdf <- Zdf %>%
  left_join(
    weather_df,
    by = c("STAT_KODE", "WEATHER_HOUR"), relationship = "many-to-many"
  )
```

#Create variable for the ARS_GRP_KODE at the previous record
```{r}
Xdf <- Xdf %>%
  arrange(JOURNEY_ID, PLANTID_datetime) %>%
  group_by(JOURNEY_ID) %>%
  mutate(
    PREVIOUS_ARS_GRP_KODE =
      lag(ARS_GRP_KODE) %>%
      coalesce("NONE") %>%
      factor()
  ) %>%
  ungroup()
```

#Create cancelled variable for every record
```{r}
Xdf <- Xdf %>%
  mutate(
    CANCELLED = factor(
      if_else(MLD_KAT_KODE %in% c("A"), "1", "0"),
      levels = c("0", "1")
    )
  )
#Checking the proportions
table(Xdf$CANCELLED)
prop.table(table(Xdf$CANCELLED))
```

#Create variable indicating if the train was cancelled at the previous station
```{r}
Xdf <- Xdf %>%
  group_by(JOURNEY_ID) %>%
  arrange(PLANTID) %>%
  mutate(
    PREVIOUS_CANCELLED =
      lag(
        CANCELLED,
        default = factor("0", levels = levels(CANCELLED))
      )
  ) %>%
  ungroup()
```

#Create previous delay variable
```{r}
Xdf <- Xdf %>%
  arrange(JOURNEY_ID, PLANTID_datetime) %>%
  group_by(JOURNEY_ID) %>%
  mutate(
    PREVIOUS_DELAY =
      lag(PLANAFVIGELSE)
  ) %>%
  ungroup()
```


#Make sure NAs match no. of journeys 
```{r}
Xdf %>%
  summarise(
    unique_journeys = n_distinct(JOURNEY_ID),
    n_na_previous_delay = sum(is.na(PREVIOUS_DELAY)),
    difference = n_na_previous_delay - unique_journeys
  )
```

#...and set NAs to 0
```{r}
Xdf <- Xdf %>%
  mutate(
    PREVIOUS_DELAY =
      ifelse(
        is.na(PREVIOUS_DELAY),
        0,
        PREVIOUS_DELAY
      )
  )
```

#Create cumulated delay variable
```{r}
Xdf <- Xdf %>%
  arrange(JOURNEY_ID, PLANTID) %>%
  group_by(JOURNEY_ID) %>%
  mutate(
    CUMULATED_DELAY = lag(cumsum(PLANAFVIGELSE), default = 0)
  ) %>%
  ungroup()
```

#Mutate variables to factors
```{r}
Xdf <- Xdf %>%
  mutate(
    STAT_KODE = as.factor(STAT_KODE),
    PROD_KODE = as.factor(PROD_KODE),
    OD_STRK = as.factor(OD_STRK),
    DELSTR_RETNING = as.factor(DELSTR_RETNING),
    WEEKDAY = as.factor(WEEKDAY),
    CANCELLED = as.factor(CANCELLED),
    ARS_GRP_KODE = as.factor(ARS_GRP_KODE),
    MLD_TYP_KODE = as.factor(MLD_TYP_KODE),
    JOURNEY_ID = as.factor(JOURNEY_ID)
  )
```


#INSPECTION SECTION
#Planafvigelse, quantiles
```{r}
summary(Xdf$PLANAFVIGELSE)

quantile(
  Xdf$PLANAFVIGELSE,
  probs = c(
    0.90,
    0.95,
    0.975,
    0.99,
    0.995,
    0.999
  ),
  na.rm = TRUE
)
```

#Trains that are cancelled but not replaced
```{r, eval=FALSE}
Xdf %>%
  summarise(
    total_records = n(),

    cancelled_not_replaced =
      sum(
        MLD_KAT_KODE %in% c("A", "Z") &
        AFLYSNINGSKODE_EVAL !=
          "Erstattet af andet rettidigt tog",
        na.rm = TRUE
      ),

    percent_of_all_records =
      100 * cancelled_not_replaced / total_records
  )

```

#How many journeys have changes in delay?
```{r}
Xdf %>%
  group_by(JOURNEY_ID) %>%
  summarise(
    delay_changes =
      n_distinct(PLANAFVIGELSE),
    .groups = "drop"
  ) %>%
  summarise(
    journeys_with_change =
      sum(delay_changes > 1),
    total_journeys = n(),
    proportion =
      journeys_with_change / total_journeys
  )
```

#How many journeys get cancelled during the records, i.e. has partial cancellation?
```{r}
journeys_with_A_and_K <-
  Xdf %>%
  group_by(JOURNEY_ID) %>%
  summarise(
    has_A = any(MLD_KAT_KODE == "A"),
    has_K = any(MLD_KAT_KODE == "K"),
    .groups = "drop"
  ) %>%
  filter(has_A & has_K)

nrow(journeys_with_A_and_K)
```

#FILTERING
#Make list with journeys valid for the modeling. Journeys that has at least one "K" (Konstateret) is included. Check number compared to before.
```{r}
valid_journeys <- Xdf %>%
  group_by(JOURNEY_ID) %>%
  summarise(
    has_K = any(MLD_KAT_KODE == "K"),
    .groups = "drop"
  ) %>%
  filter(has_K) %>%
  pull(JOURNEY_ID)
```


#Create final df, filtering out invalid journeys defined earllier (Journeys with no "K" in MLD_KAT_KODE), and keeping only records where the train was not cancelled at the previous station
```{r}
filtered_df <- Xdf %>%
  filter(
    JOURNEY_ID %in% valid_journeys,
    PREVIOUS_CANCELLED == "0"
  )
```



#CREATE SUBSET WITH ONLY ONE RECORD PER JOURNEY
```{r}
filtered_df_one_record <- filtered_df |>
  group_by(JOURNEY_ID) |>
  group_modify(\(.x, .y) {
    if (any(.x$CANCELLED == 1, na.rm = TRUE)) {
      .x |>
        filter(CANCELLED == 1) |>
        slice_sample(n = 1)
    } else {
      .x |>
        slice_sample(n = 1)
    }
  }) |>
  ungroup()
```


#Check for specific routes that are only represented once
```{r}
filtered_df_one_record %>%
  group_by(OD_STRK) %>%
  filter(n_distinct(JOURNEY_ID) == 1) %>%
  distinct(OD_STRK)

```

#6 cases - remove these
```{r}
filtered_df_one_record <- filtered_df_one_record %>%
  group_by(OD_STRK) %>%
  filter(n_distinct(JOURNEY_ID) > 1) %>%
  ungroup()

filtered_df_one_record <- filtered_df_one_record |>
  filter(
    !is.na(HOUR_OF_DAY),
    !is.na(temperature_2m)
  )
```

# check

```{r}
filtered_df_one_record |>
  summarise(
    n_records = n(),
    n_journeys = n_distinct(JOURNEY_ID),
    n_cancelled = sum(CANCELLED == 1, na.rm = TRUE),
    cancellation_rate = mean(CANCELLED == 1, na.rm = TRUE)
  )
```

#check that no journeys are represented multiple times
```{r}
filtered_df_one_record |>
  count(JOURNEY_ID) |>
  filter(n > 1)
```


#CREATE TRAIN/TEST SPLIT WITH ONE RECORD PER JOURNEY_ID
```{r}
set.seed(150)

journey_level_one_record <- filtered_df_one_record %>%
  group_by(JOURNEY_ID) %>%
  summarise(
    journey_cancelled = factor(
      as.integer(any(CANCELLED == "1")),
      levels = c(0, 1)
    ),
    .groups = "drop"
  )

journey_split_one_record <- initial_split(
  journey_level_one_record,
  prop = 0.8,
  strata = journey_cancelled
)

train_data_one_record <- filtered_df_one_record %>%
  semi_join(
    training(journey_split_one_record) %>% select(JOURNEY_ID),
    by = "JOURNEY_ID"
  ) %>%
  mutate(JOURNEY_ID = factor(JOURNEY_ID))

test_data_one_record <- filtered_df_one_record %>%
  semi_join(
    testing(journey_split_one_record) %>% select(JOURNEY_ID),
    by = "JOURNEY_ID"
  ) %>%
  mutate(JOURNEY_ID = factor(JOURNEY_ID))
```


# cross-validation 

```{r}
# Create df for selected predictors
cv_data <- train_data_one_record %>%
  select(
    CANCELLED,
    PREVIOUS_DELAY,
    STAT_KODE,
    MLD_TYP_KODE,
    WEEKDAY,
    HOUR_OF_DAY,
    CUMULATED_DELAY,
    temperature_2m
  ) %>%
  tidyr::drop_na() %>%
  mutate(
    CANCELLED = as.numeric(as.character(CANCELLED))
  )

# Check outcome
table(cv_data$CANCELLED)
prop.table(table(cv_data$CANCELLED))

# Create stratified folds from cv_data
set.seed(150)

folds <- caret::createFolds(
  y = factor(cv_data$CANCELLED),
  k = 5,
  list = TRUE,
  returnTrain = FALSE
)

# 3. Define formulas
model_formulas <- list(
  Model_1 = CANCELLED ~ PREVIOUS_DELAY,
  
  Model_2 = CANCELLED ~ PREVIOUS_DELAY + STAT_KODE,
  
  Model_3 = CANCELLED ~ PREVIOUS_DELAY +
    STAT_KODE +
    CUMULATED_DELAY +
    MLD_TYP_KODE +
    WEEKDAY +
    HOUR_OF_DAY +
    temperature_2m
)
```


```{r}
cv_glm_metrics_folds <- function(model_formula, data, folds, model_name) {
  
  purrr::map_dfr(seq_along(folds), function(i) {
    
    valid_idx <- folds[[i]]
    
    train_fold <- data[-valid_idx, ]
    valid_fold <- data[valid_idx, ]
    
    model_fit <- glm(
      model_formula,
      data = train_fold,
      family = binomial
    )
    
    pred_prob <- predict(
      model_fit,
      newdata = valid_fold,
      type = "response"
    )
    
    eps <- 1e-15
    pred_prob <- pmin(pmax(pred_prob, eps), 1 - eps)
    
    y <- valid_fold$CANCELLED
    
    # Brier score
    brier <- mean((y - pred_prob)^2)
    
    # PR-AUC
    pr_data <- tibble(
      truth = factor(y, levels = c(1, 0)),
      .pred_1 = pred_prob
    )
    
    pr_auc_value <- yardstick::pr_auc(
      pr_data,
      truth = truth,
      .pred_1,
      event_level = "first"
    )$.estimate
    
    # McFadden's pseudo-R2 on validation fold
    ll_model <- sum(y * log(pred_prob) + (1 - y) * log(1 - pred_prob))
    
    p_null <- mean(train_fold$CANCELLED)
    p_null <- pmin(pmax(p_null, eps), 1 - eps)
    
    ll_null <- sum(y * log(p_null) + (1 - y) * log(1 - p_null))
    
    mcfadden_r2 <- 1 - (ll_model / ll_null)
    
    tibble(
      model = model_name,
      fold = i,
      mcfadden_r2 = mcfadden_r2,
      pr_auc = pr_auc_value,
      brier = brier,
      aic = AIC(model_fit)
    )
  })
}
```


```{r}
#Evaluate each model across the 5 folds 
cv_fold_results <- purrr::map_dfr(
  names(model_formulas),
  function(model_name) {
    cv_glm_metrics_folds(
      model_formula = model_formulas[[model_name]],
      data = cv_data,
      folds = folds,
      model_name = model_name
    )
  }
)

cv_fold_results
```

```{r}
#summarize cv performance across folds
cv_results2 <- cv_fold_results %>%
  group_by(model) %>%
  summarise(
    mean_mcfadden_r2 = mean(mcfadden_r2, na.rm = TRUE),
    sd_mcfadden_r2 = sd(mcfadden_r2, na.rm = TRUE),
    mean_pr_auc = mean(pr_auc, na.rm = TRUE),
    sd_pr_auc = sd(pr_auc, na.rm = TRUE),
    mean_brier = mean(brier, na.rm = TRUE),
    sd_brier = sd(brier, na.rm = TRUE),
    mean_aic = mean(aic, na.rm = TRUE),
    sd_aic = sd(aic, na.rm = TRUE),
    .groups = "drop"
  )

cv_results2
```

```{r}
#Compute baseline values
baseline_fold_results <- purrr::map_dfr(seq_along(folds), function(i) {
  
  valid_idx <- folds[[i]]
  
  train_fold <- cv_data[-valid_idx, ]
  valid_fold <- cv_data[valid_idx, ]
  
  y_valid <- valid_fold$CANCELLED
  
  eps <- 1e-15
  
  #Baseline probability = cancellation prevalence in training fold
  p_null <- mean(train_fold$CANCELLED)
  p_null <- pmin(pmax(p_null, eps), 1 - eps)
  
  # Null model for AIC baseline
  null_model <- glm(
    CANCELLED ~ 1,
    data = train_fold,
    family = binomial
  )
  
  tibble(
    fold = i,
    pr_auc_baseline = mean(y_valid),
    brier_baseline = mean((y_valid - p_null)^2),
    mcfadden_r2_baseline = 0
  )
})

baseline_summary <- baseline_fold_results %>%
  summarise(
    pr_auc_baseline = mean(pr_auc_baseline, na.rm = TRUE),
    brier_baseline = mean(brier_baseline, na.rm = TRUE),
    mcfadden_r2_baseline = mean(mcfadden_r2_baseline, na.rm = TRUE)
  )

baseline_summary
```

#Summary plot, cross-validated evaluation metrics
```{r}
metric_colors <- c(
  "AIC<br><span style='font-size:9pt'>(lower is better)</span>" = "#E68613",          
  "Brier score<br><span style='font-size:9pt'>(lower is better)</span>" = "#2C7FB8", 
  "McFadden's pseudo-R²<br><span style='font-size:9pt'>(higher is better)</span>" = "#7B3294", 
  "PR-AUC<br><span style='font-size:9pt'>(higher is better)</span>" = "#D33682"      
)

#Prepare plot data
plot_data <- cv_fold_results %>%
  pivot_longer(
    cols = c(mcfadden_r2, pr_auc, brier, aic),
    names_to = "metric_raw",
    values_to = "value"
  ) %>%
  mutate(
    metric = case_when(
      metric_raw == "mcfadden_r2" ~ "McFadden's pseudo-R²<br><span style='font-size:9pt'>(higher is better)</span>",
      metric_raw == "pr_auc" ~ "PR-AUC<br><span style='font-size:9pt'>(higher is better)</span>",
      metric_raw == "brier" ~ "Brier score<br><span style='font-size:9pt'>(lower is better)</span>",
      metric_raw == "aic" ~ "AIC<br><span style='font-size:9pt'>(lower is better)</span>",
      TRUE ~ metric_raw
    ),
    model = case_when(
      model == "Model_1" ~ "Model 1",
      model == "Model_2" ~ "Model 2",
      model == "Model_3" ~ "Model 3",
      TRUE ~ model
    ),
    model = factor(model, levels = c("Model 1", "Model 2", "Model 3")),
    metric = factor(metric, levels = names(metric_colors))
  )

#Baselines
baseline_data <- tibble(
  metric_raw = c("mcfadden_r2", "pr_auc", "brier"),
  baseline = c(
    baseline_summary$mcfadden_r2_baseline,
    baseline_summary$pr_auc_baseline,
    baseline_summary$brier_baseline
  )
) %>%
  mutate(
    metric = case_when(
      metric_raw == "mcfadden_r2" ~ "McFadden's pseudo-R²<br><span style='font-size:9pt'>(higher is better)</span>",
      metric_raw == "pr_auc" ~ "PR-AUC<br><span style='font-size:9pt'>(higher is better)</span>",
      metric_raw == "brier" ~ "Brier score<br><span style='font-size:9pt'>(lower is better)</span>",
      TRUE ~ metric_raw
    ),
    metric = factor(metric, levels = names(metric_colors))
  )

# Plot
cv_plot <- ggplot(plot_data, aes(x = model, y = value, color = metric)) +
  geom_hline(
    data = baseline_data,
    aes(yintercept = baseline, color = metric),
    linetype = "dashed",
    linewidth = 0.55,
    alpha = 0.7,
    inherit.aes = FALSE
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 2.1
  ) +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.14,
    linewidth = 0.6
  ) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_color_manual(values = metric_colors, guide = "none") +
  labs(
    x = "GLM specification",
    y = "Metric Value",
    title = "Cross-validated comparison of GLM specifications",
    subtitle = "Points show mean performance across folds; error bars show ±1 SD; dashed lines show relevant baselines"
  ) +
  theme_minimal(base_size = 12) +
theme(
  plot.title = element_text(face = "bold", size = 14),
  plot.subtitle = element_text(size = 10),
  
  strip.text = ggtext::element_markdown(
    face = "bold",
    size = 11,
    margin = margin(t = 3, r = 3, b = 1, l = 3)
  ),
  
  panel.border = element_rect(
    color = "grey70",
    fill = NA,
    linewidth = 0.5
  ),
  panel.spacing = unit(1.2, "lines"),
  
  panel.grid.major.y = element_line(
    color = "grey88",
    linewidth = 0.35
  ),
  panel.grid.minor.y = element_line(
    color = "grey94",
    linewidth = 0.25
  ),
  panel.grid.major.x = element_line(
    color = "grey92",
    linewidth = 0.25
  ),
  panel.grid.minor.x = element_blank(),
  
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 9.5),
  axis.title = element_text(size = 11)
)

cv_plot
```



#GVIF for model 3

```{r}
# Model 3
model_3 <- glm(
  CANCELLED ~ PREVIOUS_DELAY +
    STAT_KODE +
    CUMULATED_DELAY +
    MLD_TYP_KODE +
    WEEKDAY +
    HOUR_OF_DAY +
    temperature_2m,
  data = train_data_one_record,
  family = binomial
)

vif(model_3)
```



#UPDATED EVALUATION FOR MODEL 3

```{r}
summary(model_3)

# McFadden's pseudo-R² for model_3

null_model_3 <- glm(
  CANCELLED ~ 1,
  data = train_data_one_record,
  family = binomial()
)

mcfadden_r2_model_3 <- 1 - as.numeric(logLik(model_3) / logLik(null_model_3))

mcfadden_r2_model_3

# Predict on held-out test set

test_data_one_record <- test_data_one_record %>%
  mutate(
    pred_prob = predict(
      model_3,
      newdata = test_data_one_record,
      type = "response"
    ),
    CANCELLED = factor(CANCELLED, levels = c("0", "1")),
    y_numeric = as.numeric(CANCELLED == "1")
  )


# Test-set performance metrics


pr_auc_test_model_3 <- pr_auc_vec(
  truth = test_data_one_record$CANCELLED,
  estimate = test_data_one_record$pred_prob,
  event_level = "second"
)


brier_test_model_3 <- mean(
  (test_data_one_record$pred_prob - test_data_one_record$y_numeric)^2
)

baseline_prob <- mean(train_data_one_record$CANCELLED == "1")

brier_baseline_test <- mean(
  (baseline_prob - test_data_one_record$y_numeric)^2
)

model_3_test_metrics <- tibble(
  model = "Model 3",
  McFadden_R2 = mcfadden_r2_model_3,
  PR_AUC = pr_auc_test_model_3,
  Brier_score = brier_test_model_3,
  Brier_baseline = brier_baseline_test
)

model_3_test_metrics


#  drop1 likelihood-ratio test for model_3


drop1_model_3 <- drop1(
  model_3,
  test = "Chisq"
)

drop1_model_3
#seed was not set for creating the filtered_df_one_record which might create slightly different results when re-running
```


#Calibration plot, 20 bins
```{r}
model_col <- "#0072B2"

test_data_one_record %>%
  mutate(pred_bin = ntile(pred_prob, 20)) %>%
  group_by(pred_bin) %>%
  summarise(
    mean_predicted = mean(pred_prob, na.rm = TRUE),
    observed_rate = mean(CANCELLED == "1", na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = mean_predicted, y = observed_rate)) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    linewidth = 0.8,
    colour = "grey55"
  ) +
  geom_line(
    colour = model_col,
    linewidth = 1.2
  ) +
 geom_point(
  colour = model_col,
  fill = "white",
  shape = 21,
  size = 1.5,
  stroke = 1.2,
  alpha = 0.9
) +
  scale_x_continuous(
    trans = pseudo_log_trans(sigma = 0.003),
    labels = percent_format(accuracy = 0.1)
  ) +
  scale_y_continuous(
    trans = pseudo_log_trans(sigma = 0.003),
    labels = percent_format(accuracy = 0.1)
  ) +
  scale_size_continuous(
    range = c(2.5, 6),
    guide = "none"
  ) +
  labs(
    x = "Mean predicted cancellation probability",
    y = "Observed cancellation rate",
    title = "Calibration plot for Model 3",
    subtitle = "Predictions are grouped into 20 equally sized bins, represented by the dots",
    caption = "Dashed line indicates perfect calibration"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, colour = "grey35"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

#seed was not set for creating the filtered_df_one_record which might create slightly different results when re-running

```


#Quantiles probability for each class
```{r}
quantile(
  test_data_one_record$pred_prob[test_data_one_record$CANCELLED == 0],
  probs = c(.5, .9, .95, .99),
  na.rm = TRUE
)

quantile(
  test_data_one_record$pred_prob[test_data_one_record$CANCELLED == 1],
  probs = c(.5, .9, .95, .99),
  na.rm = TRUE
)
```


#Visualization of predicted probabilities for outcome class, quantiles
```{r}
quantile_plot_data <- test_data_one_record %>%
  mutate(
    outcome = if_else(
      as.character(CANCELLED) == "1",
      "Cancelled",
      "Not cancelled"
    )
  ) %>%
  group_by(outcome) %>%
  summarise(
    Median = quantile(pred_prob, probs = 0.50, na.rm = TRUE),
    `90th percentile` = quantile(pred_prob, probs = 0.90, na.rm = TRUE),
    `95th percentile` = quantile(pred_prob, probs = 0.95, na.rm = TRUE),
    `99th percentile` = quantile(pred_prob, probs = 0.99, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -outcome,
    names_to = "quantile",
    values_to = "predicted_probability"
  ) %>%
  mutate(
    quantile = factor(
      quantile,
      levels = c(
        "Median",
        "90th percentile",
        "95th percentile",
        "99th percentile"
      )
    )
  )

#plot
ggplot(
  quantile_plot_data,
  aes(
    x = quantile,
    y = predicted_probability,
    group = outcome,
    colour = outcome
  )
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(
    values = c(
      "Cancelled" = "#7B2CBF",      
      "Not cancelled" = "#E69F00"   
    )
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 0.35),
    breaks = seq(0, 0.35, by = 0.05)
  ) +
  labs(
    title = "Quantile plot: Distribution of Predicted Cancellation Probabilities",
    x = NULL,
    y = "Predicted cancellation probability",
    colour = "Observed outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    axis.title.y = element_text(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```



